Hippocampal replay of experience at slow speeds

### 2021-03-10 Article.tex
Update on Overleaf.

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[4]{Marielena Sosa}
\author[5]{Jason E. Chung}
\author[6]{Uri T. Eden}
\author[1, 2, 3]{Loren M. Frank}

@@

% Figure 1
\begin{figure}
\includegraphics[width=0.75\linewidth]{figures/Figure1/Figure1_v6}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}}\label{figsupp:fig1sup1}
\end{figure}

@@

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \citep{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We first confirmed that our decoding algorithm could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal (with the caveat that theta sweeps and replay sequences will represent positions other than where the animal is and our resolution is limited by the size of our 3 cm position bins). We found, similar to other studies, that the average median difference is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation), but importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during run and we were accurately able to decode immobility times as well, demonstrating the power of our algorithm.

Having established the accuracy of our decoder, we next wanted to assess the overall prevalence of spatial content across SWRs, so we detected SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events. As expected, and providing further support for the validity of the model, we observe many replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using the standard decoder (\FIGSUPP[2]{fig2sup2}A-F). However, we also observe many events with spatially coherent content that do not have this structure. For example, we observe trajectories that start in one direction and reverse back to the original position (\FIG{3}A, \FIGSUPP[3]{fig3sup1}B), trajectories that remain fixed in one position (\FIG{3}B, \FIGSUPP[3]{fig3sup1}G), and trajectories that jump between arms and between dynamics (\FIG{3}C, D, F, \FIGSUPP[3]{fig3sup1}F, H, I). We also observe SWRs that are spatially incoherent throughout the SWR (\FIGSUPP[3]{fig3sup1}A, D).

Using a 0.8 threshold, we are able to classify 89\% (23,071 of 25,844) of SWRs as containing at least one of the three dynamics or common dynamic mixtures. To ensure that this depends on the sequential firing of spatially tuned cells in the hippocampus, we resample position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decode the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup2}).

@@

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, but instead correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (14,989 of 21,433 or 70\% of classified SWRs) are well described by a single dynamic or mixtures of dynamics (\FIG{5}A). Surprisingly, the most common category is stationary-continuous mixtures (8,944 of 21,433 or 42\% of classified SWRs, \FIG{5}A, \FIGSUPP[5]{fig5sup1}). These events contain representations that move at slower speeds (\FIG{5}D), and are slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 73 ms vs. continuous 94 ms, \FIG{5}B). Nonetheless, both these slow-moving events and continuous events frequently represent locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 52 cm vs. continuous 43 cm, \FIG{5}C). This indicates that the content of these SWRs, like those that are classified as continuous, do not represent the animal's actual position.

@@

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (100 cm/s), which is consistent with the running speed of rats (citation?). The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated an extended representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events (\texttildelow50-100 ms) with faster continuous dynamics, consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics could also represent either the animal's current location or locations far from the animal, and similarly, previous work has shown that events with continuous dynamics most often begin at the animal's current location \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

These results further challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead, there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, a previous paper \citep{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We believe that the difference in these findings is a result of defining "chance" levels of occurrence of these events and using a restrictive decoding model which requires the decoded position to be stationary for the entire event. We also note that we chose to include a stationary dynamic to explicitly evaluate the prevalence of stationary events in adult animals, but in principle, a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds close to 0 m/s.

@@

Our results showed that 19\% of SWRs contained continuous trajectories---meaning that portions of the SWR had spatially coherent trajectories that moved at high velocities---but only 4\% were continuous throughout the entire SWR. This, \textit{prima facie}, may seem at odds with previous results, which found between 5\% and 45\% of SWRs were "significant", meaning that they were well fit by a line compared to various shuffled versions of the dataset. Furthermore, these studies found trajectories on average move at much higher speeds (20x normal rat movement speed) than we report here. While the plurality of methods and parameters makes it hard to directly compare, we believe there are several reasons for this difference. We use a model with smaller time bins that allows for changes of speeds within SWRs, whereas methods using a linear fit use larger time bins and cannot accommodate rapid change. Studies using a linear fit typically incorporate position bins within 15 cm of the best fit line in their best fit score in order to compensate for the rigidness of this restriction, but this allows for a wide range of speeds when using a 20 ms time bin (up to 3.75 m/s). The linear fit also depends on the start and end event boundaries. Many studies explicitly excluded slow events.

>>>>

% Use letters for affiliations, numbers to show equal authorship (if applicable) and to indicate the corresponding author
\author[1, 2, 3]{Eric L. Denovellis}
\author[2, 3]{Anna K. Gillespie}
\author[2, 3]{Michael E. Coulter}
\author[4]{Marielena Sosa}
\author[5]{Jason E. Chung}
\author[6]{Uri T. Eden}
\author[1, 2, 3]{Loren M. Frank}

@@

% Figure 1
\begin{figure}
\includegraphics[width=0.75\linewidth]{figures/Figure1/Figure1_v7}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time, which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). Red dotted line in the right panel is the best fit line from the standard decoder. \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}}\label{figsupp:fig1sup1}
\end{figure}

@@

After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \citep{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We confirmed that our decoding algorithm could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal (with the caveat that theta sweeps and replay sequences will represent positions other than where the animal is and our resolution is limited by the size of our 3 cm position bins). We found, similar to other studies, that the average median difference is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation), but importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during run and we were accurately able to decode immobility times as well, demonstrating the power of our algorithm.

Having established the accuracy of our decoder, we next wanted to assess the overall prevalence of spatial content across SWRs. We detected SWRs using a permissive threshold (see Methods) to ensure that we include both the rare large amplitude events as well as the much more common small amplitude events. As expected, and providing further support for the validity of the model, we observe many replays that are classified as continuous throughout the entirety of the SWR, similar to those seen using the standard decoder (\FIGSUPP[2]{fig2sup2}A-F). However, we also observe many events with spatially coherent content that do not have this structure. For example, we observe trajectories that start in one direction and reverse back to the original position (\FIG{3}A, \FIGSUPP[3]{fig3sup1}B), trajectories that remain fixed in one position (\FIG{3}B, \FIGSUPP[3]{fig3sup1}G), and trajectories that jump between arms and between dynamics (\FIG{3}C, D, F, \FIGSUPP[3]{fig3sup1}F, H, I). We also observe SWRs that are spatially incoherent throughout the SWR (\FIGSUPP[3]{fig3sup1}A, D).

Using a 0.8 threshold, we are able to classify 89\% (23,071 of 25,844) of SWRs as containing at least one of the three dynamics or common dynamic mixtures. To ensure that this depends on the firing of spatially tuned cells in the hippocampus, we resample position with replacement for two recording sessions, shuffling the relationship between spiking and spatial information and then fitting the encoding model. We then decode the same SWR events containing the same spikes. Only 9\% of the SWRs are classified in the shuffled datasets, a value that is significantly less than that seen for the real data (p=0.02 for recording session 1, p=0.02 for recording session 2, \FIGSUPP[3]{fig3sup2}).

@@

Surprisingly, our results indicate that most events with coherent spatial content are not dominated by the continuous movement dynamic, but instead correspond to trajectories that are stationary or that move relatively slowly compared to their purely continuous counterparts. We therefore examined these events in more detail. We first note that most of the SWRs (14,989 of 21,433 or 70\% of classified SWRs) are well described by a single dynamic or mixtures of dynamics (\FIG{5}A). Surprisingly, the most common category is stationary-continuous mixtures (8,944 of 21,433 or 42\% of classified SWRs, \FIG{5}A, \FIGSUPP[5]{fig5sup1}). These events contain representations that move at slower speeds (\FIG{5}D), and are sustained, but slightly shorter, on average, than events with continuous dynamics (median duration: stationary-continuous-mixture 73 ms vs. continuous 94 ms, \FIG{5}B). Nonetheless, both these slow-moving events and continuous events frequently represent locations that were some distance away from the animal (mean trajectory distance from animal's position: stationary-continuous-mixture 52 cm vs. continuous 43 cm, \FIG{5}C). This indicates that the content of these SWRs, like those that are classified as continuous, do not represent the animal's actual position. We note that roughly a third of these events persisted for the entire duration of the SWR, but the other two thirds had some portion of time where the model was uncertain about the dynamic. This shows the power of our classifier because it allows us to characterize the content of a SWR even if the model is only certain about a portion of the SWR time.

@@

We performed several controls to see if our results were influenced by the firing of interneurons, which typically do not project to other brain areas and therefore cannot be read by a downstream region. First, we checked if requiring five or more tetrodes to have spikes during the SWR (up from requiring two tetrodes be active in the main results) would influence the proportions of dynamics, with the thought that a single tetrode with a high firing interneuron might result in stationary or stationary-continuous mixture dynamics during SWRs. This had little effect on the overall proportions, with many stationary and stationary-continuous mixtures still present (\FIGSUPP[5]{fig5sup2}A). Indeed, most of our events included spikes from upwards of 10 tetrodes (\FIGSUPP[5]{fig5sup3}B), including within each dynamic (\FIGSUPP[5]{fig5sup3}A). Next, we tried removing spikes that had spike widths of less than 0.3 ms because interneurons typically have narrow spike widths. Note that this is a rather imprecise threshold, because typically this is computed on the average waveform for a putative cell, not on a single spike, and combined with a requirement for the putative cell to have a high firing rate, as interneurons typically exhibit high firing rates. Nevertheless, even with this imprecise criteria, we still observed a high proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}B). We also looked at the 9 of 10 animals in our dataset that already had clustered cells, excluding putative interneurons based on spike width and firing rate and requiring that at least three putative pyramidal cells be active during the SWR. Although there were many fewer events to examine, we still observed many stationary and stationary-continuoux mixtures in SWRs compared to continuous sequences (\FIGSUPP[5]{fig5sup2}D). This provides further evidence that our results were not the result of the firing of interneurons.

Another possible concern is that many of these slow spatially coherent events occur when the animal is moving very slowly, meaning the stationary-continuous mixtures potentially reflect a slow theta sweep rather than a replay event. We do not believe this to be the case because the majority of our events occur when the animal is moving at speeds < 2 cm/s (\FIGSUPP[5]{fig5sup3}B) and when we restrict ourselves to SWRs where the animal was moving at speeds < 1 cm/s, we find similar proportion of SWRs with stationary and stationary-continuous mixture dynamics (\FIGSUPP[5]{fig5sup2}C).

@@

Importantly, the spatially coherent events most often expressed a dynamic that was consistent with slower speeds of less than 1 m/s (<100 cm/s), which is consistent with the running speed of rats. The next most common category were stationary events \citep{YuDistincthippocampalcorticalmemory2017, FarooqEmergencepreconfiguredplastic2019} that activated an extended representation of a single location, while the "classical" replay events---continuous, extended trajectories through space---made up only about 19\% of classified events. Events with slower dynamics tended to persist for approximately the same length of time as events (\texttildelow50-100 ms) with faster continuous dynamics, consistent with a brief activation of a representation of a location or a short trajectory through space. Events with slower dynamics could also represent either the animal's current location or locations far from the animal, and similarly, previous work has shown that events with continuous dynamics most often begin at the animal's current location \citep{DavidsonHippocampalReplayExtended2009, KarlssonAwakereplayremote2009}.

These results challenge the long-held notion that hippocampal replay trajectories necessarily proceed at many times the speed of the animal \citep{NadasdyReplayTimeCompression1999, LeeMemorySequentialExperience2002, DavidsonHippocampalReplayExtended2009}. Instead, there is a much richer variety of speeds the animal can use to retrieve information about prior experience or generate possible future events, including slower moving representations that activate a spatially compact set of nearby locations and stationary representations of single locations. Interestingly, a previous paper \citep{FarooqEmergencepreconfiguredplastic2019} reported that these stationary representations were most prevalent in juvenile rats and decreased to "chance" levels when the rats reached adulthood. Here, we show that they do not disappear, but are a very common occurrence with adult rats. We believe that the difference in these findings is a result of defining "chance" levels of occurrence of these events and using a restrictive decoding model which requires the decoded position to be stationary for the entire event. We also note that we chose to include a stationary dynamic to explicitly evaluate the prevalence of stationary events in adult animals, but in principle, a model with just continuous and fragmented dynamics could capture these events as well, as the continuous dynamic includes movement speeds close to 0 m/s.

@@

Our results showed that 19\% of SWRs contained continuous trajectories---meaning that portions of the SWR had spatially coherent trajectories that moved at high velocities---but only 4\% were continuous throughout the entire SWR. This, \textit{prima facie}, may seem at odds with previous results, which found between 5\% and 45\% of SWRs were "significant", meaning that they were well fit by a line compared to various shuffled versions of the dataset. Furthermore, these studies found trajectories on average move at much higher speeds (20x normal rat movement speed) than we report here. While the plurality of methods and parameters makes it hard to directly compare, we believe there are several reasons for this difference. First, our model is much more flexible because it allows the trajectory to change speed during the SWR, whereas fitting a line assumes a constant speed trajectory. So many of the events that had continuous dynamics, and would be classified as having high speed trajectories, also exhibited stationary-continuous mixture dynamics (see for example \FIG{1}). Second, the line has to be fit to the whole SWR event, meaning that it is sensitive to the estimation of the start end of the SWR event. In contrast, our model makes an estimate for each time point and only depends on the previous and future time step, which minimizes the dependence of the event on those boundaries and any possible "noise" events within the SWR.

We use a model with smaller time bins that allows for changes of speeds within SWRs, whereas methods using a linear fit use larger time bins and cannot accommodate rapid change. Studies using a linear fit typically incorporate position bins within 15 cm of the best fit line in their best fit score in order to compensate for the rigidness of this restriction, but this allows for a wide range of speeds when using a 20 ms time bin (up to 3.75 m/s). The linear fit also depends on the start and end event boundaries. Many studies explicitly excluded slow events.
